Servis: A dependently typed DSL for web APIs

### 2016-08-22 paper/paper.tex
more text

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{minted}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\author{Arian van Putten \\ a.vanputten@uu.nl}
\title{Servis: A dependently typed DSL for web APIs}
\begin{document}
\maketitle
\begin{abstract}
In this paper, we describe a DSL for web APIs named Servis. Servis is heavily inspired by Servant, a haskell type-level DSL for web APIs,  but tries to solve identified shortcomings by using a dependently typed language called Idris. Servis uses the universe pattern to deduce types from an API description such that we can verify at compile time that the implementation of the API matches the description.

Furthermore, we take a look at if dependent types can bring more to the table in terms of expressiveness of API descriptions than the original Servant DSL\@. Specifically, we introduce $\pi$-types to describe interdependencies within an API description.

  TODO: rewrite abstract \emph{after} I finish the paper :-)

@@

\section{Introduction}
Servant is an extensible type-level DSL for Haskell, that tries to make web APIs \emph{first-class}~\cite{servant}. A web API should be a thing that can be named, passed to, or returned by functions. Just like any other term in a language.

@@

TODO give a short overview of servant

@@

Dependent Haskell is currently in the making, and probably planned for around GHC version 8.6~\cite{neeeded}. In the future, it would probably be very feasible to port this library back to Haskell and bring all the benefits along with it.

@@

Preferably, one would want to define a new open kind \mintinline{haskell}{API} such that \mintinline{haskell}{(:>) :: API -> API -> API}, but this is not possible in Haskell (or Idris). Instead, one could define a closed kind (using \textt{XDataKinds}), in which we define the entire Servant DSL. This would mean we would lose extensibility of the DSL, but we gain kind-safety.

@@

@@

It would be nice to add a list of possible status codes that an endpoint can returns, and then also be able to check whether a server implementation actually only returns one of the specified status codes if something goes wrong (or right).

\subsection{Overlapping routes}
Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
It would be nice if we could verify at compile time that no routes in an API specification can overlap.

@@

\section{Servis}
In this section, I present a DSL in Idris named Servis. It tries to address each issue listed in Section~\ref{sec:issues}. Firstly, we will define the DSL in which we can write API specifications. Then, we will look at how to write an interpreter of this DSL that generates documentation. Then, we will look at how we would implement HTTP webserver handlers that adhere to specification defined by the API DSL\@. Next, we'll look at solving the overlapping routes problem by writing proofs in Idris.
Finally, we will extend the DSL to add \emph{pi-types}, allowing us to describe dependencies in our API\@.  For example, one might want to describe  the size of a list of users in the API, but let the size depend on a query parameter \emph{limit} that limits the response size.  It would then force the implementor of the API to only return values smaller or equal to limit in size.

@@

--  el api evaluates to  Int -> IO User
handler : el api
handler = getUser
 \end{minted}
\subsection{The DSL}
So what does this type \mintinline{idris}{API} look like?
\subsubsection{Handler}\label{sec:handler}
Before we will look at routes, lets think of what an endpoint should look like.
We should be able to handle the HTTP verbs. We will only implement \mintinline{idris}{GET} and \mintinline{idris}{POST}, because the others are very similar.
For a \mintinline{idris}{GET}, we should be able to specify a return type of the handler,  and for a \mintinline{idris}{POST}, we should be able to
specify a request body type, and a return type.

@@

||| PathPart is a part of a path.
||| @ capture the universe of captures
||| @ query   the universe of query params
data PathPart : (capture : Type) -> (query : Type) -> Type where
  ||| A constant piece of text
  ||| @ path the path part
  Const : (path : String) -> PathPart capture query
  ||| A capture of a variable name of a specific type
  ||| @ name  the name of the variable
  ||| @ type  the type of the variable
  Capture : (name : String) -> (type : capture) -> PathPart capture query
  ||| A query param of a variable name of a specific type
  ||| @ name  the name of the variable
  ||| @ type  the type of the variable
  QueryParam : (name : String) -> (type : query) -> PathPart capture query
\end{minted}

A \mintinline{idris}{Path} is simply a list if \mintinline{idris}{PathPart}s, which ends in a \mintinline{idris}{Handler}:
\begin{minted}{idris}
||| Describes a Path. A Path consists of path parts followed by a handler
data Path : (capture : Type) -> 
            (query : Type) ->
            (req : Type) ->
            (res : Type) -> Type where
  ||| Ends a path with a handler
  ||| @ handler the handler
  Outputs : (handler : Handler req res) -> Path capture query req res
  ||| Conses a pathpart to a path
  (:>) : PathPart capture query ->
         Path capture query req res ->
         Path capture query req res
infixr 5 :>
\end{minted}

@@

findUserByName : String -> IO (List User)
getUsers : (limit : Int) -> IO (List User)
getUserById : (id : Int) -> IO User
addUser : User -> IO ()
\end{minted}
\label{domain}
\caption{The User object and its interactions}
\end{listing}

@@

\begin{listing}
\begin{minted}{idris}
data QueryUniverse
  = QSTRING
  | QINT
Universe QueryUniverse where
  el QSTRING = String
  el QINT = Int

@@

userAPI : API QueryUniverse CaptureUniverse ReqResUniverse ReqResUniverse
userAPI = OneOf
  [ Const "users" :> Capture "user_id" CINT :> Outputs (GET USER)
  , Const "users" :>
    Const "find" :>
    QueryParam "email" QSTRING :>
    Outputs (GET (LIST USER))
  , Const "users" :> Outputs (POST USER)
  , Const "users" :>
    QueryParam "limit" QINT :>
    Outputs (GET (LIST USER))
  ]
\end{minted}
\label{code:api}
\caption{API definition for users}
\end{listing}

@@

# /users?limit=<QINT>
## GET
 Response type: LIST USER

@@

# /users?limit=<QINT>
## GET
 Response type LIST USER
\end{verbatim}

@@

\subsection{A proof of non-overlapping routes}
We will describe types for helping prove that routes in an API definition don't overlap.  We do this
by providing Decidable Equality proofs for all parts of our DSL.
\begin{minted}{idris}

\end{minted}
\subsection{Pi-types}
Adds dependent pairs to our DSL such that we can describe dependencies between query params, captures and return types.  Example:

\begin{minted}{haskell}
exampleOfDependentPath : Path EmptyU QueryU EmptyU RespU
exampleOfDependentPath =
  QueryParam "limit" NAT :*> \limit => Outputs (GET (BLIST limit USER))

getUsersLimit : (limit : Nat) -> IO (BoundedList limit User)
getUsersLimit limit = pure . take limit $ users

exampleOfDependentPathHandler : el Example.exampleOfDependentPath
exampleOfDependentPathHandler = getUsersLimit

\end{minted}
\subsection{New problems that pi-type create}
Here we will describe how pi-types adds problems. For example we cannot easily generate documentation anymore, prove that there are no overlapping routes, or prove that statuscodes are adhered to.

\section{Summary}
\section{Related work}
Talk about tel/serv. An inspiration for this project.
\section{Conclusions and future work}

\bibliography{sources}
\bibliographystyle{plain}
\end{document}

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{minted}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\author{Arian van Putten \\ a.vanputten@uu.nl}
\title{Servis: A dependently typed DSL for web APIs}
\begin{document}
\maketitle
\begin{abstract}
\emph{TODO}
%In this paper, we describe a DSL for web APIs named Servis. Servis is heavily inspired by Servant, a haskell type-level DSL for web APIs,  but tries to solve identified shortcomings by using a dependently typed language called Idris. Servis uses the universe pattern to deduce types from an API description such that we can verify at compile time that the implementation of the API matches the description.

%Furthermore, we take a look at if dependent types can bring more to the table in terms of expressiveness of API descriptions than the original Servant DSL\@. Specifically, we introduce Sigma-types to describe interdependencies within an API description.

% TODO: rewrite abstract \emph{after} I finish the paper :-)

@@

\section{Introduction}
\emph{TODO expand and reword}

Servant is an extensible type-level DSL for Haskell, that tries to make web APIs \emph{first-class}~\cite{servant}. A web API should be a thing that can be named, passed to, or returned by functions. Just like any other term in a language.

@@

\emph{TODO give a short overview of servant}

@@

Dependent Haskell is currently in the making, and probably planned for around GHC version 8.6. In the future, it would probably be very feasible to port this library back to Haskell and bring all the benefits along with it.

@@

Preferably, one would want to define a new open kind \mintinline{haskell}{API} such that \mintinline{haskell}{(:>) :: API -> API -> API}, but this is not possible in Haskell (or Idris). Instead, one could define a closed kind (using \texttt{XDataKinds}), in which we define the entire Servant DSL. This would mean we would lose extensibility of the DSL, but we gain kind-safety.

@@

\subsection{API Interdependencies}
\label{sec:inter}
Sometimes, one one like to describe even more type-safety than currently possible in Servant. Maybe we want to let the output-type of a request depend on a query parameter. A use case could be the following: We have a route \texttt{"/users/list?limit=100}", and we want to make sure that a handler never returns more than 100 users.  Thus the return type of the handler should be a \mintinline{idris}{BoundedList 100 User}. Or more generally, the return type should be \mintinline{idris}{BoundedList limit User}, where \mintinline{idris}{limit : Nat} comes from parsing the URL.  The full type of the API will thus depend on some runtime value. This is a textbook example of dependent types, and is currently not possible to implement in Haskell. However, with idris this should be a breeze.

\subsection{Overlapping routes}
Another problem with Servant is that there is no protection for overlapping routes. If two routes could resolve to the same path, Servant makes a left-biased choice~\cite{servant}.
It would be nice if we could verify at compile time that no routes in an API specification can overlap and give the user an error if he violates the overlapping constraint.

@@

It would be nice to add a list of possible status codes that an endpoint can returns, and then possibly also be able to check whether a server implementation actually only returns one of the specified status codes if something goes wrong (or right).

@@

\section{Servis}
In this section, I present a DSL in Idris named Servis. It tries to address each issue listed in Section~\ref{sec:issues}. Firstly, we will define the DSL in which we can write API specifications. Then, we will look at how to write an interpreter of this DSL that generates documentation. Then, we will look at how we would implement HTTP webserver handlers that adhere to specification defined by the API DSL\@. Next, we'll look at solving the overlapping routes problem by using Idris's automatic proof search.
Finally, we will extend the DSL to add dependent types, allowing us to describe dependencies in our API\@.  For example, one might want to describe  the size of a list of users in the API, but let the size depend on a query parameter \emph{limit} that limits the response size.  It would then force the implementor of the API to only return values smaller or equal to limit in size.

@@

--  el api evaluates to  Int -> IO User
handler : el api
handler = getUser
 \end{minted}
\subsection{The DSL}
So what does this type \mintinline{idris}{API} look like? An API DSL should allow us to define routes, to which we can attach handler types.
\subsubsection{Handler}\label{sec:handler}
Before we will look at routes, lets think of what an endpoint should look like.
We should be able to handle the HTTP verbs. We will only implement \mintinline{idris}{GET} and \mintinline{idris}{POST}, because the others are very similar.
For a \mintinline{idris}{GET}, we should be able to specify a return type of the handler,  and for a \mintinline{idris}{POST}, we should be able to
specify a request body type, and a return type.

@@

||| PathPart is a part of a path.
||| @ capture the universe of captures
||| @ query   the universe of query params
data PathPart : (capture : Type) -> (query : Type) -> Type where
  ||| A constant piece of text
  ||| @ path the path part
  Const : (path : String) -> PathPart capture query
  ||| A capture of a variable name of a specific type
  ||| @ name  the name of the variable
  ||| @ type  the type of the variable
  Capture : (name : String) -> (type : capture) -> PathPart capture query
  ||| A query param of a variable name of a specific type
  ||| @ name  the name of the variable
  ||| @ type  the type of the variable
  QueryParam : (name : String) -> (type : query) -> PathPart capture query

( Universe capture
, Universe query
) => Universe (PathPart capture query) where
  el (Const path) = ()
  el (Capture name type) = el type
  el (QueryParam name type) = el type
\end{minted}

A \mintinline{idris}{Path} is simply a list if \mintinline{idris}{PathPart}s, which ends in a \mintinline{idris}{Handler}:
\begin{minted}{idris}
||| Describes a Path. A Path consists of path parts followed by a handler
data Path : (capture : Type) -> 
            (query : Type) ->
            (req : Type) ->
            (res : Type) -> Type where
  ||| Ends a path with a handler
  ||| @ handler the handler
  Outputs : (handler : Handler req res) -> Path capture query req res
  ||| Conses a pathpart to a path
  (:>) : PathPart capture query ->
         Path capture query req res ->
         Path capture query req res
infixr 5 :>

( Universe capture
, Universe query
, Universe req
, Universe resp
) => Universe (Path capture query req resp) where
  -- special case because we don't like () in our functions
  el (Const path :> right) =  el right
  el (pathPart :> right) = el pathPart -> el right
  el (Outputs handler) = el handler
\end{minted}

@@

findUserByEmail : String -> IO (List User)
getUsers : (limit : Nat) -> IO (List User)
getUserById : (id : Int) -> IO User
addUser : User -> IO User
\end{minted}
\label{code:domain}
\caption{The User object and its interactions}
\end{listing}

@@

\begin{listing}
\begin{minted}{idris}
data QueryUniverse
  = QSTRING
  | QINT
  | QNAT
Universe QueryUniverse where
  el QSTRING = String
  el QNAT = Nat

@@

userAPI : API QueryUniverse CaptureUniverse ReqResUniverse ReqResUniverse
userAPI = OneOf
  [ Const "users" :> Capture "user_id" CINT :> Outputs (GET USER)
  , Const "users" :>
    Const "find" :>
    QueryParam "email" QSTRING :>
    Outputs (GET (LIST USER))
  , Const "users" :> Outputs (POST USER)
  , Const "users" :>
    QueryParam "limit" QNAT :>
    Outputs (GET (LIST USER))
  ]

-- set the handlers
handlers : el userAPI
handlers =
  [ getUserById
  , findUserByEmail
  , addUser
  , getUsers
  ]

\end{minted}
\label{code:api}
\caption{API definition for users}
\end{listing}

@@

# /users?limit=<QNAT>
## GET
 Response type: LIST USER

@@

\end{verbatim}

@@

Now we will look at how we will route a \mintinline{idris}{Path}. As we know, a path 
is a description of how we extract information from the URL. Because parsing URLs
is out of scope, we will assume that URLs are already parsed when they are fed into our
routing machinery as the following datatype:
\begin{minted}{idris}
record URL where
  constructor MkURL
  pathParts : List String
  params : List (String, String)
\end{minted}

\clearpage
We will now give a step-by-step description of the path router.
\begin{minted}{idris}
( FromCapture capture
, FromQueryParam query
, FromRequest req
, ToResponse resp
) => HasServer (Path capture query req resp) where
  -- routing a Const is simple. Just make sure the const matches
  -- the next path part
  route ((Const path) :> right)
        handler
        (MkURL pathParts params) 
        requestBody =
    case pathParts of
      -- if there are no pathParts left, abort. There is no match!
      [] => Nothing
      (x::xs) => do
        -- make sure the pathPart we consume matches the path part we expect
        guard (x == path)
        --- recurse on the remaining path parts
        route right handler (MkURL xs params) requestBody

  -- To route a capture, we pass the captured variable along to our handler
  route ((Capture name type) :> right)
        handler
        (MkURL pathParts params)
        requestBody =
    case pathParts of
      -- if there are no pathParts left, abort. There is no match!
      [] => Nothing
      (x::xs) => do
        -- try to parse the pathPart as a capture parameter of the right type
        -- abort if this fials
        capture <- fromCapture type x
        -- pass the parsed capture to handler, and recurse on
        -- the other pathparts
        route right (handler capture) (MkURL xs params) requestBody
\end{minted}
\pagebreak
\begin{minted}{idris}
  -- To route a query param, we look up the query param in the URL,
  -- and pass it to the handler
  route ((QueryParam name type) :> right)
        handler
        (MkURL pathParts params)
        requestBody = do
    -- make sure the query param that was expected exists
    val <- lookup name params
    -- try parse it to the appropriate type
    param <- fromQueryParam type val
    -- remove the parsed query param from the URL
    let newParams = dropWhile ((== name) . fst) params
    -- apply the param to the handler,
    -- and recurse on the url without the query param
    route right (handler param) (MkURL pathParts newParams) requestBody
\end{minted}

Routing multiple paths is easy. We just keep on trying consecutive paths until one
doesn't return \mintinline{idris}{Nothing}:

\begin{minted}{idris}

( FromCapture capture
, FromQueryParam query
, FromRequest req
, ToResponse resp) =>
  HasServer (API capture query req resp) where

  route (OneOf (path :: []))
        (handler :: [])
        url
        requestBody =
    -- if no handlers are left but this one.
    -- We should allow it to fail
    route path handler url requestBody
  route (OneOf (path :: x :: xs))
        (handler :: h :: handlers)
        url
        requestBody =
    -- if other handlers are present. Try the handler
    -- and otherwise try others
    case route path handler url requestBody of
      Nothing =>
        route (OneOf (x :: xs))
              -- NOTE: this is to help the compiler select correct type
              (the (HVect _ ) (h :: handlers)
              url
              requestBody
      a => a
\end{minted}
\subsection{A proof of non-overlapping routes}
\emph{TODO}
\subsection{Adding API interdependencies}
As discussed in Section~\ref{sec:inter}, it would be nice to describe more complexly typed APIs, where we allow interdependencies within an API path.

We can easily add this functionality to our DSL by introducing a \emph{dependent} version of \mintinline{idris}{:>}, named
\mintinline{idris}{:*>} in the \mintinline{idris}{Path} datatype.
What we want is that that the value of the lefthand-side of \mintinline{idris}{:*>} influences the type of the right-hand side.
\begin{minted}{idris}
data Path : capture -> query -> req -> res -> Type where
  Outputs : (handler : Handler req res) -> Path capture query req res
  (:>) :  (pathPart : PathPart capture query) ->
          (path : Path capture query req res) -> Path capture query req res
  -- The RHS of :*> now takes a function instead of a path. that calculates a path
  (:*>) : (pathPart : PathPart capture query) ->
          (path : (el pathPart -> Path capture query req res)) ->
          Path capture query req res
\end{minted}{idris}
This is enough to express basic dependent APIs.

A good usecase for these kind of dependent types is for endpoints that return a \mintinline{idris}{BoundedList}.
A bounded list in idris is a list that can only contain upto a certain amount of elements. It is defined as follows:
\begin{minted}{idris}
data BoundedList : Nat -> Type -> Type where
  Nil : BoundedList n a
  (::) : a -> BoundedList n a -> BoundedList (S n) a
\end{minted}

Lets extend our Universe of Listing~\ref{code:api} to include bounded lists:
\begin{minted}{idris}
data ReqResUniverse
  = USER
  | LIST ReqResUniverse
  | BLIST Nat ReqResUniverse
Universe ReqResUniverse where
  el USER = User
  el (LIST a) = List (el a)
  el (BLIST n a) = BoundedList n (el a)
\end{minted}
And lets modify the domain function that queries the database for users
in Listing~\ref{code:domain} to return bounded lists:

\begin{minted}{idris}
getUsers : (limit : Nat) -> IO (BoundedList limit User)
\end{minted}
We can now describe an API in which we can describe the dependency between the query parameter
and the size of the returned list of users.
\begin{minted}{idris}
getUsersPath : Path CaptureUniverse QueryUniverse ReqResUniverse ReqResUniverse
getUsersPath = Const "users" :> 
              QueryParam "limit" QNAT :*>
              \limit => Outputs (GET (BLIST limit USER))

handler : el getUsersPath
handler = getUsers
\end{minted}

Next, we will need to add extra cases to the \mintinline{idris}{HasServer}
implementation of \mintinline{idris}{Path} to handle the dependency. We
will only show the implementation for \mintinline{idris}{QueryParam},
but the changes to make \mintinline{idris}{Capture} work are identical.

The idea is, that instead of only passing the parsed query parameter value
to the handler, we also pass it along to the right hand side of \mintinline{idris}{:*>}:
\begin{minted}{idris}
  route ((QueryParam name type) :*> right)
        handler
        (MkURL pathParts params)
        requestBody = do
    val <- lookup name params
    param <- fromQueryParam type val
    let newParams = dropWhile ((== name) . fst) params
    -- BEFORE :
--  route right 
--        (handler param)
--        (MkURL pathParts newParams)
--         requestBody
    -- WITH ADDED DEPENDENCY:
    route (right param) -- <- note the extra argument passed to RHS
          (handler param)
          (MKURL pathParts newParams)
          requestBody
\end{minted}
And that's it. we now have added dependent types to our API!

\subsection{New problems that dependent types create}
The dependent types are nice, but our datatype is now a Higher-order abstract syntax instead of a first-order abstract syntax.
Which means the datatype contains a function.  This makes it difficult to generate documentation as we need to provide a value
to apply to the function to traverse the datatype. It isn't impossible though. We could introduce the following interface:
\begin{minted}{idris}
interface Universe u => HasSample u where
  sample : (v : u) -> el v

data ExampleU = INT | STRING

Universe ExampleU where
  el INT = Int
  el STRING = String

HasSample ExampleU where
  sample INT = 0
  sample STRING = "A string"
\end{minted}
Now in our implementation of \mintinline{idris}{HasDocs} we can just call \mintinline{idris}{sample}
every time we encounter a function:
\begin{minted}{idris}
HasDocs (Path capture query req res) where
  docs ((Capture str type) :*> f) =
    "super nice docs"  ++ docs (f (sample type))
\end{minted}

The real problem though is that this representation breaks the overlapping path detector.
The overlapping path detector uses the feature of idris to search proofs automatically. But that
proof search isn't perfect, and won't be able to prove that paths overlap or not if we have
arbitrary functions in our datatypes.

A solution this problem would be that library authors would have to provide their own proof
of whether or not their api definitions overlap, but this can be very cumbersome and thus
is undesired.

\section{Related work}
This project was heavily inspired by \url{https://github.com/tel/serv}. Which is
a closed-kinded Servant alternative. When I stumbled upon it, I got very curious
what would happen if we would port it to Idris.  The serv source code is kind of
dense because Haskell isn't a very nice language to do type-level programming in
and it got me curious whether or not idris would improve readability. Which it did!

Also, once I had a working example of Serv in idris, I wanted to see if idris
could add any new tricks to it. This made me implement dependent types and
route conflict resolution.

\section{Conclusions and future work}
Idris allows us to elegantly describe APIs with plain ol' datatypes. We
can then use the Universe pattern to map these datatypes to handler
function types. These handler function types can then be made routable
by writing an interpreter of the API datatype that decodes the incoming
requests and passes the correct parameters to the handlers.

Idris makes it easier to write utility functions, like documentation generation
or code generation, because our API definition does not live in the type-level
anymore, because type-level and term-level are united in idris.

We were able to solve the common mistake of overlapping routes in Servant as well,
by using Idris's proof search capabilities.

Finally, we were able to implement an dependent API, that allows interdependencies
within paths to be able to describe even more powerful APIs.

In the future, it would be interesting to look at if it's possible to enforce in 
the type-system that handlers only return HTTP Status codes that are defined
in the API description. I explored some ideas on how to implement this, but
did not have the time to do an implementation of this.
\bibliography{sources}
\bibliographystyle{plain}
\end{document}
